Estimating initial state and quark-gluon plasma medium properties \\ using a hybrid model with nucleon substructure \\ calibrated to p-Pb and Pb-Pb collisions at \texorpdfstring{$\mathbf{\sqrts=5.02}$}{}~TeV

### 2018-08-01 paper/substructure.tex
Start finishing summary and conclusion

@@

\newcommand{\trento}{T\raisebox{-0.5ex}{R}ENTo}
\newcommand{\sqrts}{\sqrt{s_{NN}}}
\newcommand{\nch}{N_\text{ch}}
\newcommand{\ntrk}{N_\text{trk}^\text{offline}}
\newcommand{\sigmaf}{\sigma_\mathrm{fluct}}
\newcommand{\X}{\chi_\mathrm{struct}}
\newcommand{\taufs}{\tau_\mathrm{fs}}
\newcommand{\dmin}{d_\mathrm{min}}
\newcommand{\T}{\tilde{T}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\z}{\mathbf{z}}
\newcommand{\vnk}[2]{v_#1\{#2\}}
\newcommand{\paddedhline}{\noalign{\smallskip}\hline\noalign{\smallskip}}
\newcommand{\order}[1]{$\mathcal O(10^{#1})$}
\newcommand{\note}{\textcolor{theblue}{[?]}}

@@

In this work, we pursue a less restrictive and more parametric description of the proton where the number of substructure degrees of freedom are uncertain as depicted in Fig.~\ref{fig:substructure}.
We model each nucleon's density $\rho_{A,B}$ as a sum of $n_c$ independent constituents
\begin{equation}
  \label{eq:rho}
  \rho_{A, B}(\x) = \frac{1}{n_c} \sum\limits_{i=1}^{n_c} \rho_c(\mathbf{x} - \mathbf{x_i}),
\end{equation}
where each constituent density $\rho_c$ is described by a Gaussian distribution of width $v$
\begin{equation}
  \label{eq:constituent_density}
  \rho_c(\mathbf{x}) = \frac{1}{(2\pi v^2)^{3/2}} \exp\left(-\frac{\x^2}{2 v^2}\right),
\end{equation}
and each constituent's position $\mathbf{x_i}$ in Eqn.~\eqref{eq:rho} is sampled from a Gaussian radial distribution with standard deviation $r$.

@@

\begin{figure}
  \begin{tikzpicture}
    \node {\includegraphics{thickness}};
    \node[label=above:{10 fm}] (a) at (0, 1.1) {};
    \node (b) at (0, -1.1) {};
    \draw [<->, semithick] (a) to (b);
  \end{tikzpicture}
  \caption{Effect of nucleon substructure on the nuclear thickness function $T(x, y) \equiv \int dz\, \rho(x, y, z)$ of a $^{208}\mathrm{Pb}$ nucleus. The nucleus on the left has Gaussian nucleons of width $0.8$~fm, while the nucleus on the right has composite nucleons, each containing six constituents of width $0.4$~fm.}
  \label{fig:thickness}
\end{figure}

@@

We now direct our attention to Fig.~\ref{fig:posterior} which shows the main result of this work, the posterior distribution of the model input parameters.
Recall that the posterior $P(H_\x | E)$ is the probability that our hypothesis $H_\x$ for the true model parameters $\x_\star$ is correct, given the evidence $E$ provided by experiment.
The present posterior has 15 dimensions, one for each parameter listed in Tbl.~\ref{tab:design}, and thus its joint distribution cannot be summarized by one figure alone.
We thus sample the distribution and histogram the samples to project the distribution onto one or two dimensions at a time.
Each diagonal panel is the distribution of a \emph{single} model parameter (marginalized over all others), and each lower-diagonal panel is the joint distribution of a \emph{pair} of model parameters (marginalized over all others).

The posterior in Fig.~\ref{fig:posterior} includes a wealth of information.
Here we focus on its most important features, starting with the top-left corner of the figure and working our way to the bottom-right.
We report posterior estimates for individual parameters using their (marginalized) distribution medians and 90\% credible intervals.
For example, the estimate $x=2.45_{-0.15}^{+0.20}$ denotes a median value $\tilde{x}=2.45$ and 90\% credible interval $2.30 < x < 2.65$.

@@

@@

@@

@@

\begin{figure}[t]
  \includegraphics{posterior_parton_number}
  \caption{\label{fig:posterior_parton_number}
    Posterior distribution for the number of nucleon constituents $n_c$ determined by the analysis.
    The parameter $n_c$ is an integer (discrete) variable at every design point, but the emulator interpolation produces a posterior distribution which is continuous.
  }
\end{figure}

@@

\begin{figure}
  \includegraphics{posterior_proton_shape}
  \caption{\label{fig:posterior_proton_shape}
  Posterior distribution for the nucleon width $w$ and constituent width $v$. The white trapezoidal region (lower-right) covers the initial range of allowed values for $w$ and $v$, while the gray triangular region (upper-left) indicates values which are excluded by the prior $v > w$. The posterior distribution, shown in blue, indicates the preferred values for $w$ and $v$ determined by the analysis.}
\end{figure}

@@

\begin{table}[b]
  \caption{
    \label{tab:mode_params}
    High-probability parameters selected from the posterior distribution and used to generate Fig.~\ref{fig:obs_map}.
    The posterior distribution on the particlization temperature $T_\text{switch}$ is flat (agnostic), so we fix it's value using Ref.~\cite{Bernhard:2018hnz}.
  }
  \begin{ruledtabular}
    \begin{tabular}{ll@{\hspace{2em}}ll}
      \multicolumn{2}{c}{Initial condition} & \multicolumn{2}{c}{QGP medium} \\
      \paddedhline
      Norm     & 20.            & $\eta/s$ min      & 0.11           \\
      $p$      & 0.0            & $\eta/s$ slope    & 1.6 GeV$^{-1}$ \\
      $k$      & 0.19           & $\eta/s$ curv     & -0.29          \\
      $n_c$    & 6              & $\zeta/s$ max     & 0.032          \\
      $w$      & 0.92 fm        & $\zeta/s$ width   & 0.024 GeV      \\
      $v$      & 0.43 fm        & $\zeta/s$ $T_0$   & 175 MeV        \\
      $\dmin$  & 0.81 fm        & $T_\text{switch}$ & 151 MeV        \\
      $\taufs$ & 0.37 fm/$c$
    \end{tabular}
  \end{ruledtabular}
\end{table}

@@

@@

We built, for this purpose, a flexible multi-stage nuclear collision model characterized by a number of free parameters which vary theoretically uncertain aspects of the framework such as the QGP initial conditions and hydrodynamic transport properties.
For the QGP initial conditions, we developed a modified version of the \trento\ model which adds new parameters to vary the fluctuating size and shape of each nucleon.
Specifically, we modeled each nucleon as a cluster of $n_c$ constituents (hot spots), where each constituent is described by a Gaussian density profile of width $v$.
We also introduced a parameter $w$ which effectively varies the width of the Gaussian distribution used to sample constituent positions inside each nucleon.

The transport dynamics of the collision were simulated using a pre-equilibrium free-streaming stage of variable duration followed by boost-invariant viscous hydrodynamics for hot and dense regions of the fireball and a microscopic hadronic afterburner for the dilute corona.
We parametrized various sources of uncertainty in each stage of the collision including the duration of the pre-equilibrium free-streaming stage, the temperature dependence of the QGP shear and bulk viscosities, and the isothermal particlization temperature used to switch from a hydrodynamic description to microscopic Boltzmann transport.

We applied Bayesian parameter estimation which were developed to estimate the parameters of computationally expensive model.

\medskip

...work in progress

\vfill

@@

\begin{acknowledgments}
  JSM thanks Weiyao Ke, Bj\"orn Schenke and Heikki M\"antysaari for helpful discussions and clarifying comments.
  This research was completed using 6 million CPU hours provided by the National Energy Research Scientific Computing Center (NERSC), a U.S. Department of Energy Office of Science User Facility operated under Contract No. DE-AC02-05CH11231.
  JSM, JEB and SAB acknowledge support by ...
\end{acknowledgments}

@@

\begin{figure}
  \begin{tikzpicture}
    \draw [black, thin, step=0.3cm] (0.9, 0.9) grid +(4.5, 4.5);
    \draw [gray, thin, step=0.9cm] (0, 0) grid +(6.3, 6.3);
    \draw [theblue] (3.15, 3.15) circle (2.25cm);
  \end{tikzpicture}
  \caption{Diagram of the adaptive grid resizing algorithm (not drawn to scale). Each initial condition event is first run on a very large coarse-grained mesh (large gray grid) of one-third the spatial resolution otherwise required to measure hydrodynamic observables. We then measure the maximum transverse radius $R_\text{max}$ (blue circle) of the hypersurface defined by the temperature isotherm $T = T(e_\text{min})$, where $e_\text{min}$ is the largest energy density which can be truncated without modifying the hydrodynamic observables calculated from the event. Finally, the initial condition event is rerun on a smaller and finer mesh (smaller black grid) with three-times the cell density of the pre-run event and a smaller transverse extent $-R_\text{max} < x < R_\text{max}$.}
  \label{fig:adaptive_grid}
\end{figure}

@@

>>>>

\usepackage{booktabs}

@@

\newcommand{\trento}{T\raisebox{-0.5ex}{R}ENTo}
\newcommand{\sqrts}{\sqrt{s_{NN}}}
\newcommand{\nch}{N_\text{ch}}
\newcommand{\ntrk}{N_\text{trk}^\text{offline}}
\newcommand{\Tsw}{T_\text{switch}}
\newcommand{\sigmaf}{\sigma_\mathrm{fluct}}
\newcommand{\X}{\chi_\mathrm{struct}}
\newcommand{\taufs}{\tau_\mathrm{fs}}
\newcommand{\dmin}{d_\mathrm{min}}
\newcommand{\fmc}{\ensuremath{\text{fm}/c}}
\newcommand{\T}{\tilde{T}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\z}{\mathbf{z}}
\newcommand{\vnk}[2]{v_#1\{#2\}}
\newcommand{\paddedhline}{\noalign{\smallskip}\hline\noalign{\smallskip}}
\newcommand{\order}[1]{$\mathcal O(10^{#1})$}
\newcommand{\note}{\textcolor{theblue}{[?]}}

@@

In this work, we pursue a less restrictive and more parametric description of the proton where the number of substructure degrees of freedom are uncertain as depicted in Fig.~\ref{fig:substructure}.
We model each nucleon's density $\rho_{A,B}$ as a sum of $n_c$ independent constituents
\begin{equation}
  \label{eq:rho}
  \rho_{A, B}(\x) = \frac{1}{n_c} \sum\limits_{i=1}^{n_c} \rho_c(\mathbf{x} - \mathbf{x_i}),
\end{equation}
where each constituent density $\rho_c$ is described by a Gaussian distribution of width $v$
\begin{equation}
  \label{eq:constituent_density}
  \rho_c(\mathbf{x}) = \frac{1}{(2\pi v^2)^{3/2}} \exp\left(-\frac{\x^2}{2 v^2}\right),
\end{equation}
and each constituent's position $\mathbf{x_i}$ in Eqn.~\eqref{eq:rho} is sampled from a Gaussian radial distribution with standard deviation $r$.
The effect of this addition nucleon substructure on the nuclear thickness functions is visible in Fig.~\ref{fig:thickness}

@@

\begin{figure}
  \begin{tikzpicture}
    \node {\includegraphics{thickness}};
    \node[label=above:{10 fm}] (a) at (0, 1.1) {};
    \node (b) at (0, -1.1) {};
    \draw [<->, semithick] (a) to (b);
  \end{tikzpicture}
  \caption{\label{fig:thickness} Effect of nucleon substructure on the nuclear thickness function $T(x, y) \equiv \int dz\, \rho(x, y, z)$ of a $^{208}\mathrm{Pb}$ nucleus. The nucleus on the left has Gaussian nucleons of width $0.8$~fm, while the nucleus on the right has composite nucleons, each containing six constituents of width $0.4$~fm.}
\end{figure}

@@

We now direct our attention to Fig.~\ref{fig:posterior} which shows the main result of this work, the posterior distribution of the model input parameters.
Recall that the posterior $P(H_\x | E)$ is the probability that our hypothesis $H_\x$ for the true model parameters $\x_\star$ is correct, given the evidence $E$ provided by experiment.
The present posterior has 15 dimensions, one for each parameter listed in Tbl.~\ref{tab:design}, and thus its joint distribution cannot be summarized by one figure alone.
We thus sample the distribution and histogram the samples to project the distribution onto one or two dimensions at a time.
Each diagonal panel is the distribution of a \emph{single} model parameter (marginalized over all others), and each lower-diagonal panel is the joint distribution of a \emph{pair} of model parameters (marginalized over all others).
We also report numeric estimates for each model parameter along the distribution diagonal.
These estimates describe the marginalized distribution median and 90\% credible interval for each parameter.
For example, the estimate $x=2.45_{-0.15}^{+0.20}$ denotes a median value $\tilde{x}=2.45$ and 90\% credible interval $2.30 < x < 2.65$.
These posterior parameter estimates are also listed in Tbl.~\ref{tab:post_param} for clarity.

@@

\begin{table}[t]
  \caption{
    \label{tab:post_param}
    Posterior parameter estimates corresponding to Fig.~\ref{fig:posterior}.
    The reported values are for the distribution median and 90\% highest posterior density credible interval.
  }
  \begin{ruledtabular}
    \newlength{\cellwidth}
    \settowidth{\cellwidth}{$-0.00$}
    \newcommand{\est}[3]{\parbox{\cellwidth}{\hfill$#1$}$_{-#2}^{+#3}$}
    \begin{tabular}{llll}
      \toprule
      \multicolumn{2}{c}{Initial condition / Pre-eq}     & \multicolumn{2}{c}{QGP medium}              \\
      \cmidrule(r){1-2}                                    \cmidrule(l){3-4}
      \addlinespace[.4ex]
      Norm       & \est{20.3}{2.2}{2.2}          & $\eta/s$ min     & \est{0.10}{0.07}{0.07}            \\[1.1ex]
      $p$        & \est{0.018}{0.159}{0.146}     & $\eta/s$ slope   & \est{1.16}{1.16}{1.55} GeV$^{-1}$ \\[1.1ex]
      $\sigmaf$  & \est{0.93}{0.27}{0.27}        & $\eta/s$ crv     & \est{-0.21}{0.79}{0.85}           \\[1.1ex]
      $w$        & \est{0.97}{0.17}{0.19} fm     & $\zeta/s$ max    & \est{0.027}{0.027}{0.034}         \\[1.1ex]
      $n_c$      & \est{6.2}{3.1}{2.8}           & $\zeta/s$ width  & \est{0.032}{0.032}{0.047} GeV     \\[1.1ex]
      $\X$       & \est{0.34}{0.17}{0.22}        & $\zeta/s$ $T_0$  & \est{0.176}{0.020}{0.024} GeV     \\[1.1ex]
      $\dmin$    & \est{1.12}{0.50}{0.56} fm     & $\Tsw$           & \est{0.150}{0.012}{0.015} GeV     \\[1.1ex]
      $\taufs$   & \est{0.37}{0.27}{0.34} \fmc   & \\
      \addlinespace[.4ex]
      \bottomrule
    \end{tabular}
  \end{ruledtabular}
\end{table}

@@

\begin{figure}[b]
  \includegraphics{posterior_parton_number}
  \caption{\label{fig:posterior_parton_number}
    Posterior distribution for the number of nucleon constituents $n_c$ determined by the analysis.
    The parameter $n_c$ is an integer (discrete) variable at every design point, but the emulator interpolation produces a posterior distribution which is continuous.
  }
\end{figure}

@@

\begin{figure}
  \includegraphics{posterior_proton_shape}
  \caption{\label{fig:posterior_proton_shape}
  Posterior distribution for the nucleon width $w$ and constituent width $v$. The enclosed trapezoidal region covers the prior range of allowed values for $w$ and $v$. The posterior distribution, shown in blue, indicates the preferred values for $w$ and $v$ determined by the analysis.}
\end{figure}

@@

@@

@@

@@

\begin{table}
  \caption{
    \label{tab:mode_params}
    High-probability parameters selected from the posterior distribution and used to generate Fig.~\ref{fig:obs_map}.
    The posterior distribution on the particlization temperature $T_\text{switch}$ is flat (agnostic), so we fix it's value using Ref.~\cite{Bernhard:2018hnz}.
  }
  \begin{ruledtabular}
    \begin{tabular}{ll@{\hspace{2em}}ll}
      \multicolumn{2}{c}{Initial condition / Pre-eq} & \multicolumn{2}{c}{QGP medium} \\
      \paddedhline
      Norm     & 20.            & $\eta/s$ min      & 0.11           \\
      $p$      & 0.0            & $\eta/s$ slope    & 1.6 GeV$^{-1}$ \\
      $k$      & 0.19           & $\eta/s$ curv     & -0.29          \\
      $n_c$    & 6              & $\zeta/s$ max     & 0.032          \\
      $w$      & 0.92 fm        & $\zeta/s$ width   & 0.024 GeV      \\
      $v$      & 0.43 fm        & $\zeta/s$ $T_0$   & 175 MeV        \\
      $\dmin$  & 0.81 fm        & $T_\text{switch}$ & 151 MeV        \\
      $\taufs$ & 0.37 fm/$c$
    \end{tabular}
  \end{ruledtabular}
\end{table}

@@

We built, for this purpose, a flexible multi-stage nuclear collision model characterized by a number of free parameters which vary theoretically uncertain aspects of the framework such as the QGP initial conditions and hydrodynamic transport properties.
For the QGP initial conditions, we employed a modified version of the \trento\ model which adds new parameters to vary the fluctuating size and shape of each nucleon.
Specifically, we modeled each nucleon as a cluster of $n_c$ constituents (hot spots), where each constituent is described by a Gaussian density profile of width $v$.
Each constituent position inside the nucleon was sampled randomly (without correlations) according to a Gaussian radial distribution of variable width, controlled indirectly by an effective nucleon width parameter $w$.

The transport dynamics of the collision were simulated using a pre-equilibrium free-streaming stage of variable duration followed by boost-invariant viscous hydrodynamics for hot and dense regions of the fireball and a microscopic hadronic afterburner for the relatively dilute corona.
We parametrized various sources of uncertainty in each stage of the collision including the duration of the pre-equilibrium free-streaming stage, the temperature dependence of the QGP shear and bulk viscosities, and the particlization temperature used to switch from a hydrodynamic description to microscopic Boltzmann transport.

With the full evolution model in hand, we applied Bayesian methods which were developed to estimate the parameters of computationally intensive models \note.
We first constructed a scaffolding of $n=500$ parameters points distributed throughout our 15-dimensional parameter space and evaluated the nuclear collision model using \order{4} events at each parameter point.
The ensemble of events was then used to calculate a large number of experimental observables at each design point and train Gaussian process emulators to interpolate the model predictions as a function of the input parameters.
Finally, we used Markov chain Monte Carlo (MCMC) importance sampling to explore the parameter space and draw samples from the Bayesian posterior distribution for the \emph{true} values of the model parameters, given our emulated model predictions, the experimental data, and their associated uncertainties.

The model calibration process is summarized by Figs.~\ref{fig:obs_pbpb} and \ref{fig:obs_ppb}, and the resulting posterior distribution for the model input parameters is shown in Fig.~\ref{fig:posterior}.
We also evaluated the model predictions using a single set of high-probability parameters in Fig.~\ref{fig:obs_map}.
With these results, we are able to address the two primary goals of the study.

First, we demonstrated in Fig.~\ref{fig:obs_map} the existence of a single set of model parameters which can simultaneously describe $p$-Pb and Pb-Pb charged-particle yields, mean $p_T$, and flow cumulants at $\sqrts=5.02$~TeV.
The excellent quantitative agreement of the model calculation with the experimental data is strong evidence for a unified hydrodynamic description of $p$-A and A-A collisions.
Moreover, the modifications to the physics model which were required to obtain this agreement are generally modest; one must simply replace Gaussian nucleons with composite nucleons of several or more constituents.

Second, we obtained a posterior distribution for the model input parameters in Fig.~\ref{fig:posterior} and reported quantitative estimates for the model input parameters in Tbl.~\ref{tab:post_param}.
Here we summarize our key findings about the model:
\begin{enumerate}[itemsep=0pt, leftmargin=2\parindent]
  \item
    Using the \trento\ initial condition model, we find that initial entropy deposition (or particle production) scales like the geometric mean of local participant nuclear densities, in agreement with a number of previous estimates \note.
  \item
    Nucleon substructure is necessary to simultaneously describe $p$-Pb and Pb-Pb bulk observables. However, there is no strong preference for a specific number of constituents inside the nucleon.
  \item
    Bulk observables in $p$-Pb and Pb-Pb collisions prefer larger nucleons, similar in size to the proton's RMS electric charge radius.
    This agrees with a similar Bayesian analysis which was calibrated to Pb-Pb bulk observables at $\sqrts=2.76$ and 5.02~TeV using a hybrid model without nucleon substructure \note.
    We note that our result is in tension an estimate for the effective nucleon width based on the gluon distribution extracted from HERA data \note.
  \item
    We obtain an estimate $v=0.46_{-0.15}^{+0.21}$~fm for the Gaussian width of individual constituent hot spots inside the nucleon is which is significantly larger than the length scales typically associated with nucleon substructure.
    This result may help explain the seemingly unreasonable success of hydrodynamics in small collision systems.
  \item
    Our posterior estimate for the pre-equilibrium free-streaming time is $\taufs=0.37_{-0.27}^{+0.34}$~fm/$c$.
    This is significantly shorter than the estimate obtained from a similar Bayesian study in Ref.~\cite{Bernhard:2018hnz}, which reported $\taufs=1.16_{-0.25}^{+0.29}$~fm/$c$.
    It is not clear whether the difference is a result of nucleon substructure or the different observables used to calibrate each analysis.
  \item
    The posterior estimates for the temperature dependence of the QGP specific shear and bulk viscosities
\end{enumerate}

The present study would benefit from a number of improvements.
Perhaps the most obvious improvement would be to add missing experimental data to the calibration once it becomes available.
For example, the present calibration is missing the Pb-Pb mean $p_T$, as well as the transverse energy, identified particle yields, and mean $p_T$ fluctuations of both collision systems.
These should certainly influence the quality of the combined fit and the posterior estimates for the model parameters.

Similarly, the analysis would greatly benefit from additional beam energies and collision systems.
Notably missing are measurements of $p$-Au, $d$-Au and $^3$He-Au collisions at RHIC \note, which would provide an additional constraint on the parameters and more stringent tests of the calibrated model predictions.
The RHIC data may also help elucidate the beam-energy dependence of the model parameters which would be worth investigating.
We leave these improvements for future work.

@@

\begin{acknowledgments}
  JSM thanks Weiyao Ke, Bj\"orn Schenke and Heikki M\"antysaari for helpful discussions and clarifying comments.
  This research was completed using 6 million CPU hours provided by the National Energy Research Scientific Computing Center (NERSC), a U.S. Department of Energy Office of Science User Facility operated under Contract No. DE-AC02-05CH11231.
  JSM, JEB and SAB are supported by the U.S.\ Department of Energy Grant No.\ DE-FG02-05ER41367 and National Science Foundation Grant No.\ NSF-ACI-1550225.
\end{acknowledgments}

@@

@@

\begin{figure}[t]
  \begin{tikzpicture}
    \draw [black, thin, step=0.3cm] (0.9, 0.9) grid +(4.5, 4.5);
    \draw [gray, thin, step=0.9cm] (0, 0) grid +(6.3, 6.3);
    \draw [theblue] (3.15, 3.15) circle (2.25cm);
  \end{tikzpicture}
  \caption{Diagram of the adaptive grid resizing algorithm (not drawn to scale). Each initial condition event is first run on a very large coarse-grained mesh (large gray grid) of one-third the spatial resolution otherwise required to measure hydrodynamic observables. We then measure the maximum transverse radius $R_\text{max}$ (blue circle) of the hypersurface defined by the temperature isotherm $T = T(e_\text{min})$, where $e_\text{min}$ is the largest energy density which can be truncated without modifying the hydrodynamic observables calculated from the event. Finally, the initial condition event is rerun on a smaller and finer mesh (smaller black grid) with three-times the cell density of the pre-run event and a smaller transverse extent $-R_\text{max} < x < R_\text{max}$.}
  \label{fig:adaptive_grid}
\end{figure}
